Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2024-05-31 paper/ms.tex
more on intro and references

In particular, Scibior and Wood \cite{scibior21} show that the estimators derived in Poyiadjis et al. \cite{poyiadjis11} can be attained with AD, through with a simple tweak (the "stop-gradient trick") to the particle filter that does not modify its likelihood estimates. However, their method is not fully compatible with simulation-based inference and the mathematical motivation behind the stop-gradient trick is not apparent.

%We show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.

\kevin{Can someone who's reading through the words and glazing through the math get the idea that (1) we're running AD on the PF, and (2) using an off-policy resampling trick to get our estimator. Should I hedge by writing the SI in Del Moral language? }

\subsection{Our Contributions}
We develop a new approach to statistical inference via AD for particle filters which addresses the above concerns. We construct a new theoretical framework that bypasses the paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.

Our key observation is that one can turn this into a problem of differentiating through a series of smooth measurement density ratios and (through the reparametrization trick) a differentiable simulator. To do so, we construct a suitably reweighted family of particle filters that we call \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$), for some $\alpha \in [0,1]$ that optimizes a bias-variance tradeoff. This family encompasses the biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$. It does not require access to the transition densities, and works with just a differentiable simulator.

We show that these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties. We also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.

This allows us to develop improved algorithms for maximum likelihood/MAP estimation in highly nonlinear stochastic dynamical systems. We propose a hybrid algorithm called \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts iterative first or second-order gradient methods with a coarse solution obtained from a few rounds of iterated filtering. Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.

@@

\subsection{MOP-$\alpha$ Bias and Variance}
\begin{thm}
    \label{thm:mop-biasvar}
    When $\alpha\in(0,1)$, $\theta=\phi$, the MSE and variance of MOP-$\alpha$, where $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$, are:
    \vspace*{-1ex}
    \begin{align*}
        &\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 
        \\
        &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right),\\[-1ex]
        &\text{Var}(\nabla_\theta \hat\ell^{\alpha}(\theta)) \lesssim \min_{k\leq N} NpG'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N\right).
        \end{align*}
\end{thm}
\kevin{Explain exactly what $\epsilon$ is here, and interpret the rates as agreeing with our intuition. Talk about the relationship between the theory and the empirical results in the bias-variance tradeoff figure above, as well as the phase transition between $Nk^2/J$ and $N^4/J$ variance with MOP-$\alpha$ and MOP-$1$, where $k$ is really a surrogate for $\alpha$ -- with more memory, you truncate less. }

>>>>

Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al. \cite{poyiadjis11} can be attained with standard AD software, using an algorithmic prodecure called a ``stop-gradient'' which allows selected exressions to be evaluated but not differentiated.
Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo error.
We use \cite{scibior21} and \cite{poyiadjis11} as our starting point, while developing a new approach which remedies the weaknesses of these papers.

We start by presenting a new construction of the estimator of \cite{scibior21} and \cite{poyiadjis11}.
Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a Measurement Off-Policy (MOP) particle filter.
We procced to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$ , which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
The bias-variance tradeoff induced by $\alpha$ suggests, both in theory and in practice, the use of $\alpha$ values strictly between 0 and 1.

We then derive a linear convergence rate for stochastic gradient descent (SGD) using MOP($\alpha$) in the  presence of strong convexity.
Critically, the estimator is effective within a neighborhood of the maximum but is relatively slow at reaching this neighborhood.
That behavior is complementary to that of the iterated filtering algorithm \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
We therefore build a hybrid algorithm called \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts iterative first or second-order gradient methods with a coarse solution obtained from a few rounds of iterated filtering. Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.

@@

\subsection{MOP-$\alpha$ Bias and Variance}
\begin{thm}
    \label{thm:mop-biasvar}
    When $\alpha\in(0,1)$, $\theta=\phi$, the MSE and variance of MOP-$\alpha$, where $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$ and $\epsilon$ depends on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} are:
    \vspace*{-1ex}
    \begin{align*}
        &\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 
        \\
        &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right),\\[-1ex]
        &\text{Var}(\nabla_\theta \hat\ell^{\alpha}(\theta)) \lesssim \min_{k\leq N} NpG'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N\right).
        \end{align*}
\end{thm}
\kevin{Explain exactly what $\epsilon$ is here, and interpret the rates as agreeing with our intuition. Talk about the relationship between the theory and the empirical results in the bias-variance tradeoff figure above, as well as the phase transition between $Nk^2/J$ and $N^4/J$ variance with MOP-$\alpha$ and MOP-$1$, where $k$ is really a surrogate for $\alpha$ -- with more memory, you truncate less. }
